---
output:
  ioslides_presentation:
    widescreen: yes
    incremental: yes
---

## Writing Functions

### Writing custom functions is a great way to encapsulate code that you want to re-use


## Anatomy of a Function

### Functions have:

*  a ***name***
*  1 or more ***arguments***
*  the ***body*** of the function (i.e., code that does something with the arguments)
*  a ***return value***

## Anatomy of a Function

```{r}
isItADuck <- function(input){
  test <- input == "duck"
  return(test)
}
```

*  `isItADuck` = the name of the function
*  `input` = the only argument for this function
*   all the code between the curly brackets `{}` is the body of the function
*  we use `return()` to specify what gets returned when we call the function (in this case it will be `TRUE` or `FALSE`, because we are returning the result of a logical test)

## Anatomy of a Function


```{r}
isItADuck <- function(input){
  test <- input == "duck"
  return(test)
}
```


1.  the names `input` and `test` do not exist outside the function (i.e. after running the above code, you will get the following error if you type `input` into the R interpreter) `Error in eval(expr, envir, enclos) : object 'input' not found`

2.  Nothing happens when we create the function.  Stuff happens later when we ***call*** the function with specific arguments. 

## Asking a function to run


```{r}
isItADuck <- function(input){
  test <- input == "duck" 
  return(test)
}
```


I assume you ran the code above in R (i.e. you defined our function)

Now we can call the function with a specified value for our argument. 

```{r}
isItADuck("cat")
```

## Asking a function to run


```{r}
isItADuck("cat")
```


* When we ask `isItADuck()` to run, it needs an argument called `input`. ***But remember, this value this doesn't exist outside the function.***

* We solved this problem by providing `"cat"` as the value of `input`.



## Asking a function to run


```{r}
isItADuck("cat")
```


* Our function took an argument, determined whether or not it was a duck, and returned either `TRUE` or `FALSE`

* Now we can re-use this function over and over

## Running a function over and over

```{r}
animals <- c("cat", "dog", "duck", "chicken")
for(animal in animals){
  print(isItADuck(animal))
}
```

## Summary

### A function has arguments, that can only be referenced within the function itself. 

### When you ask a function to run, you must provide values for these arguments

### Code within the body of the function gets run

### A single value is returned