---
title: "Visualizing Data"
author: Andrew Barr
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(learnr)
library(ggplot2)
```

## base graphics in R

### 

When it comes to visualization in R, I think the base graphics in R are a bit kludgy.

We won't spend long on them, but I want you to know the basics.  

### `hist()`

You have already run into the `hist()` function in your homework.  It makes a histogram.

```{r}
hist(iris$Petal.Length)
```


### `plot()`

Appropriately enough, the major function for most kinds of plots in base R is called `plot()`.

`plot()` is a generic function, which means it can produce a variety of different types of plots depending on the kinds of variables you give it.  

### basic scatterplot

The best way to pass arguments to the plot function is using a formula, which employs the tilde character `~`.

**Example:** `plot(y~x)`: you can read this as 'plot y as a function of x'.  This means put y on the vertical axis, and x on the horizontal axis.


You will see formulas a lot once we get into doing statistical tests in R, as this is typically how you specify what kind of test you want to do. 

```{r scatter, exercise=TRUE, exercise.eval=TRUE}
plot(iris$Sepal.Width~iris$Sepal.Length)
```

### other plot types

You can produce other types of plots by specifying the `type=` argument to `plot()`.  For example, there is a built in dataset in R that consists of a time series of body temperature measurements of a beaver every 10 minutes (seriously).

```{r}
head(beaver1)
```

If we want to make a line plot, we can do the following.  Note, there are 114 temperature observations, so I am just plotting them numerically as observation 1, 2, 3, etc. Also, I am specifying `x` and `y` directly as arguments, rather than using a formula. 

```{r beaver, exercise=TRUE, exercise.eval=TRUE}
plot(x=1:114, y=beaver1$temp, type="l")
```

### basic boxplot

When the variable to the right of the formula is a factor, then `plot()` is smart enough to know that you want to plot a boxplot.

```{r box, exercise=TRUE,  exercise.eval=TRUE}
plot(iris$Sepal.Length~iris$Species)
```

### customizing plots in base R

There are a bewildering array of arguments which allow you to customize base R plots.  A few of the most common are:

* `pch` - defines the plotting symbol used (I like `pch=16`)
* `col` - the color for the plotting symbol (see the many color names R knows by typing `colors()`)
* `xlab` - title for the x axis
* `ylab` - title for the y axis
* `main` - main title for the plot


Check out `?par` for an exhaustive list of these arguments....there are A LOT!


```{r options, exercise=TRUE}
plot(iris$Sepal.Length~iris$Sepal.Width, 
     pch=16,
     col="springgreen", 
     xlab='sepal width',
     ylab='sepal length',
     main='scatterplot of sepal width and length in iris flowers')
```

## I prefer `ggplot2`

###

As I mentioned, I find these base R plots to be a bit underwhelming. 

I only showed you the how to do base R plots because you will run into them on the internet, so it helps to know the basics.  

I prefer to use the `ggplot2` package over base graphics, and I will insist you learn `ggplot2` (another `tidyverse` package).


### Dataset for tutorial

We will use a dataset called `diamonds`, which is part of the `ggplot2` package.  Lets take a look at the head (i.e. first few rows) of this dataset.

```{r warning=FALSE, message=FALSE}
head(diamonds) 
#I assume you have already done library(ggplot2)
```

## the `qplot()` function

The `ggplot2` package has a handy function called `qplot` which stands for "**q**uick **plot**". It is designed to be a shortcut to some common plots, and to work much like the standard base `plot()` function in R

### Univariate plot

The simplest `qplot()` has a variable on the x axis, and specifies a dataframe in which to look for this variable.  

`qplot()` will attempt to guess the best `geom` (graphical geometric representation) with which to display the data. 

Notice that `qplot()` correctly guesses that I want a histogram below, because this is the standard way to plot a single continuous variable.

```{r cache=TRUE}
qplot(x=carat, data=diamonds)
```

### Bivariate scatter

If we specify an additional variable for the y axis, then the default `geom` is a scatter of points.

This time we will also specify a value for `main` which is the main title. 

```{r cache=TRUE}
qplot(x=carat, y=price, data=diamonds, main="Diamond Weight by Price")
```

### Continuous versus categorical variable 

Now lets plot clarity by carat.  Note: carat is a continuous measure of weight, and clarity is a categorical variable. 

```{r cache=TRUE}
qplot(x=clarity, y=carat, data=diamonds)
```


### Boxplot 

The plot above is OK, but there are a ton of points. We can see the underlying distribution of data points better using a boxplot, by specifying `geom='boxplot'`. 

Boxplots illustrate the the median (center line), the upper and lower quartiles (the box), and show outliers as single points.

[Here is a handy reference for the available geoms.](https://ggplot2.tidyverse.org/reference/#section-geoms)

```{r cache=TRUE}
qplot(x=clarity, y=carat, data=diamonds, geom="boxplot")
```

### Layers - adding a trend line

We can add additional layers using the `+` symbol.  First, we start with the bivariate plot, then add a new layer. 

The function `stat_smooth()` adds a summary layer.

```{r cache=TRUE}
qplot(x=carat, y=price, data=diamonds, main="Diamond Weight by Price") + 
  stat_smooth()
```

### Challenge 

The relationship on the previous plot is clearly not linear.  

Make the same plot, but log transform both of the variables. When you add the trend line, use `method="lm"`, since the logged data more closely approximate a linear relationship. 

Don't save logged variables to a variable, log the variables on-the-fly when you make the plot. 

```{r logdiamond, exercise=TRUE}
### Your answer here
```

```{r logdiamond-solution }
qplot(x=log(carat), y=log(price), data=diamonds, main="Diamond Weight by Price") + 
  stat_smooth(method="lm")

```

### Color, Size, Shape, Fill {.columns-2 .build}

Visualize a third variable by changing:

>- color or fill
>- shape or size of points 

```{r cache=TRUE, fig.height=3, fig.width=5}
qplot(x=carat, y=price, 
      data=diamonds, color=clarity)
qplot(x=carat, y=price, 
      data=diamonds, shape=cut)
```

### Facets {.build}

When we have so many data points, it is hard to tell what is going on.

It would be nice to have a grid of sub-plots, so we can see the effect of these factors on the relationship between price and size. 

We can use the `facets` argument and tell it which factors we want to use, as in `facets=row_var~column_var` (here the formula notation is back, like we saw earlier in the `plot()` function)

### Facets  {.build}

```{r cache=TRUE}
qplot(x=carat, y=price, data=diamonds, 
      main="Diamond Weight by Price", facets=cut~clarity)
```




### Themes - Customize look and feel

There are several theme options that you can add to a plot, in the same way you added the `stat_smooth()` layer previously. 

You can customize almost everythign using the `theme()` function, but there are some helpful shortcut functions that I use a lot to change a bunch of theme elements at omce to get a nice effect.


### `theme_bw()`

`theme_bw()` is great because it gets rid of the default gray background. 


```{r cache=TRUE}

qplot(x=carat, y=price, data=diamonds, 
      main="Nice plot with no gray background", color=clarity) + 
    theme_bw()

```

### `theme_classic()` with BIG text

Theme classic is super basic, looking more like the base R graphics.  

You can also specify a value for the base text size in any of these themes, to increase size for presentations, etc. All the text elements will scale intelligently. Notice how much bigger all the text elements are in this plot than the previous one. 

```{r cache=TRUE}
qplot(x=carat, y=price, data=diamonds, 
      main="This theme is....classic", color=clarity) + 
  theme_classic(base_size=18)
```

## `iris` Challenge

Using the `iris` dataset and the `qplot` function, plot Sepal.Length on the x axis, and Sepal.Width on the y axis, and color code the points by Species. Style your plot using `theme_bw()`

```{r chall, exercise=TRUE}
## Your answer here
```

```{r chall-solution}
qplot(x=Sepal.Length, y=Sepal.Width, data=iris, color=Species) + 
  theme_bw()
```


## The real `ggplot()`

### 

`qplot()` is a shortcut.  You can't get all the functionality of `ggplot2` using the `qplot()` shortcut.

Lets talk about the logical structure of `ggplot2` graphics.  

### background elements

These are elements (shared by all layers in a plot)

> - `data` - a dataframe which applies to all layers. Note: `ggplot` works best when everything you need for your plot is in the same dataframe
> - `mapping` - mapping of data values into plot values (either coordinates or aesthetic properties like color values) - mappings are created by the `aes()` function
> - `facets` (optional) - specification of subplots within a plot, mapped to a column in the dataframe

### layers

Once your background is set up, you can add elements that are layered over the background.

> - `geom` - choice of how to graphically represent the data
> - `stat` - an optional statistical transformation of rows in the dataframe (sometimes this is the 'identity' tranformation, which means effectively no transformation)
> - at the layer level, you can also specify new values for `data` and `mapping` that only apply to a particular layer


## A simple `ggplot()` example

### 

Let's recreate this base plot in `ggplot2` using the car data from Motor Trends, which is available as a built in dataset called `mtcars`

```{r}
plot(mtcars$mpg~mtcars$disp, pch=16, xlab="engine size (cubic inches)", ylab="fuel efficiency (miles per gallon)", main="car fuel efficiency by engine size")
```

### background elements

First, we set up the background elements by specifying a dataframe, and mapping the columns to the different elements using `aes()`

```{r}
ggplot(data=mtcars, mapping=aes(x=disp, y=mpg))
```

We pass the dataframe through the `data` argumement, and do the mapping using the `mapping` argument. Note, when you map a column of the dataframe to a particular graphical element, you use the `aes()` function, which stands for "aesthetic".  Some people read this as "ask" as in: "ask x to represent `disp` and ask y to represent `mpg`.

Notice that we can see our axes set up, but nothing else appears because this plot doesn't have layers yet.

### layers

Next, we must specify a layer, and a geometric object `geom` to use to represent the data. Here, we add a single line of code, telling it to use the "point" geom to represent the data. 

```{r}
ggplot(data=mtcars, mapping=aes(x=disp, y=mpg)) + 
  geom_point()
```

The geoms all are in the format `geom_point`, `geom_line`, `geom_boxplot`, etc.

[Here is a handy reference for the available geoms.](https://ggplot2.tidyverse.org/reference/#section-geoms)


### change axis labels
  
Lastly, we use the `labs()` function to customize the x and y axis labels, and add a title

```{r mtcars, exercise=TRUE, exercise.eval=TRUE}
ggplot(data=mtcars, mapping=aes(x=disp, y=mpg)) + 
  geom_point() + 
  labs(x="engine size (cubic inches)", 
       y="fuel efficiency (miles per gallon)",
       title="car fuel efficiency by engine size")
```


### simple example summary

To summarize, we have set up our background using the `ggplot()` function, and mapped our aesthetics using `aes()`

Then we have added a layer, specifying which geometric object to use to represent the data. 

Finally, we have added customized labels for our axes and a title using the `labs()` function. 

## A complex `ggplot()` example

Here is a much more complex example, where we: 

* set up our background
* add `geom_boxplot()`, mapping the fill aesthetic at the level of the boxplot layer rather than the background. Note that this 
* then we add a theme
* customize our title
* set limits for our y axis using `scale_y_continuous` (I have done this to make more space at the top for my annotation)
* finally, we add a text annotation. 

```{r ggex, exercise=TRUE}
ggplot(data = diamonds, aes(x = clarity, y=price)) + 
  geom_boxplot(aes(fill=cut)) + 
  theme_bw(10) + 
  labs(title="Price by Clarity") + 
  scale_y_continuous(limits=c(0,25000)) + 
  annotate("text", x=4.5, y=23000, label="Lots of outliers", size= 4)
```


## For reference: the full `ggplot()` workflow

*   invoke the `ggplot(data=mydata, mapping=aes(x=myVar))` function to set background elements which will apply to all layers.
    * supply a base dataframe using the `data` argument
    * use `aes()` to map dataframe columns to aesthetics, using the `mapping` argument. Things you can set include
        * `x`
        * `y`
        * `color` (the color of points and lines)
        * `fill` (the color of polygons or boxplots, i.e. things that have an area to fill)
        * `shape`
        * `size` 
        * `alpha` (i.e. transparency)
        * `group` - the names of factors which speficy groups....this often set implicity when you color code by a factor.
    *   Anything not getting mapped to a column (e.g. `color="blue"`) doesn't go inside a call to `aes()`
    *   Use `facet_grid()` or `facet_wrap()` to add faceting formula. 

*   Add layers to the plot, usually using one of the `geom_xxx()` functions like:
    *  `geom_point()`
    *  `geom_histogram()`
    *  `geom_bar()`
    *  `geom_path()`, `geom_rect()`, `geom_polygon()`
    *  `geom_abline()`
    *  `geom_text()`
    *  `geom_hline()` and `geom_vline()`
    *  **Note** you can map aesthetics specific to a layer by supplying the appropriate arguments to these functions
    *  **Warning** most of these functions will automatically inherit the aesthetics from the background `ggplot()`.  If you want to override this behavior (e.g. to use a different dataframe), you can use the argument `inherit.aes=FALSE`
*   add an annotation layer `annotate()` to your plot
*   Style the plot and tweak axes etc.
    * Themes are a useful shortcut 
    * There are some helper functions for tweaking plots
      * `labs(x="The X axis label", title="my nice little graph")`
      * `guides()` to tweak legend
      * `scale_x_continuous(limits=c(0,10))` to customize x scale
      * `scale_fill_xxx()` and `scale_color_xxx()` for customizing color choices    
*   Export the plot in a useful format using `ggsave()`
*   Submit for publication.

## Final challenge

Recreate this plot yourself, using the `mtcars` dataset.

Note, the `cyl` variable is stored as a continuous variable, so you need to call `factor(cyl)` when you make the plot, to tell `ggplot()` to treat this a a categorical variable. Otherwise the fill color won't work as expected.  

```{r echo=FALSE}
ggplot(mtcars, aes(x=cyl, y=mpg, fill=factor(cyl))) +
  geom_boxplot() 
```

## Congrats!

You have finished the data visualization tutorial.
