---
title: "Spatial Data in R"
output: 
  ioslides_presentation:
    widescreen: yes

---

## Spatial data


> - Lots of data in our field(s) are inherently spatial.


> - R has lots of tools for interacting with and analyzing spatial data: 



## the `sf` package

> - The `sf` (simple features) package is a new(ish) way to interact with spatial data that simplifies dealing with spatial data in R.  


> - It simplifies the storage of spatial data into a dataframe with a special column called `geometry`. 


> - This `geometry` column allows R to plot real maps (to scale), and allows you to do lots of geospatial analyses. 


## actual spatial data versus coordinates {.smaller .build}

We often just have coordinate data, but this works differently in R from fully spatial data

```{r eval=FALSE}
myCoords <- data.frame(long = runif(20, min=35, max=36), lat = runif(20, min = 3, max=5))
plot(myCoords)
```

```{r echo=F}
set.seed(1234)
myCoords <- data.frame(long = runif(20, min=35, max=36), lat = runif(20, min = 3, max=5))
plot(myCoords)
text(x=35.7, y=4.3, labels = "What is wrong with this plot?", col="red", cex=1.4)
```

## making spatial objects {.build .smaller}


For a gold star: Which is the x and which is the y when we are dealing with latitude and longitude?

```{r}
library(sf)
SFmyCoords <- st_as_sf(myCoords, coords = c("long", "lat"))
```

## smarter plots 

```{r}
plot(SFmyCoords, axes=TRUE, xlab="longitude (degrees)", ylab="latitude (degrees)", pch=16)
```


## fully spatial object

Now there are a plethora of spatial things you can do with these points that you couldn't before

```{r}
#boundary box
st_bbox(SFmyCoords)

```

## fully spatial object

Add a buffer around the points, and then plot

```{r}
buffered <- st_buffer(SFmyCoords, 0.1)
plot(buffered, axes=TRUE, col="red", main="sites with 0.1 degree buffer")
```

## fully spatial object

Check out the `sf` cheatsheet ([here](sf.pdf)) to see the kind of spatial operations you can do 

## `rnaturalearth` {.smaller}

The `rnaturalearth` package provides maps that you can use in your plots. 

It can be a bit hard to install this package.


```{}
#try this first
install.packages(c("rnaturalearth", "rnaturalearthdata", "rnaturalearthhires"))

#if the above fails, try this
install.packages("devtools")
devtools::install_github("ropensci/rnaturalearth")
devtools::install_github("ropensci/rnaturalearthdata")
devtools::install_github("https://github.com/ropensci/rnaturalearthhires")
```

## `rnaturalearth` {.smaller}

```{r}
library(rnaturalearth)
world <- ne_countries(scale = "medium", returnclass = "sf")
plot(st_geometry(world))
```

## These objects can have lots of attributes {.smaller}

Because these are `sf` objects, they have attributes in data.frame which come in handy later for plotting

```{r}
head(world)
```


## works with `ggplot2` using `geom_sf`

```{r}
library(ggplot2)
ggplot(data=world) + 
  geom_sf(aes(fill=as.numeric(pop_est))) + 
  scale_fill_viridis_c() + 
  coord_sf(expand=F) #don't expand the map to an even number
  
```


## labeling

```{r message=F, warning=F}
countries <- ne_countries(scale="medium", country=c("Kenya", "United Republic of Tanzania", "Ethiopia", "Uganda", "Rwanda", "Burundi") ,returnclass = "sf")
ggplot(countries) + 
  geom_sf(aes(fill=sovereignt)) + geom_sf_label(aes(label=sovereignt), size=2) + guides(fill="none")

```

## more detailed maps from `rnaturalearth`

```{r}
ethiopia <- ne_states(country="Ethiopia", returnclass = "sf")
ggplot(ethiopia) + 
  geom_sf(aes(fill=name))
```

## Challenge

Use `rnaturalearth::ne_countries()` to download map data for Brazil. Plot this and make it pretty. 


## RGoogleMaps

```{r}
library("RgoogleMaps")
GW <- GetMap(center = c(38.899, -77.049), zoom = 17)
PlotOnStaticMap(GW)
```

## RGoogleMaps

```{R}
GWSatellite <- GetMap(center = c(38.899, -77.049), zoom = 17, maptype = "satellite")
PlotOnStaticMap(GWSatellite, lon=-77.0493, lat=38.899, pch=16, col="red", cex=2)
```

## GIS in R {.build}

You can do almost anything in R that you can do in ArcGIS.  

This is accomplished through additional packages, notably `rgdal` which is an R interface to the Geospatial Data Abstraction Library (GDAL), which is a great piece of open source software for spatial analysis. GDAL is the brains behind QGIS. 

The `spatstats` package also implements a variety of spatial statistics. 

One issue: R can be very slow with raster data (e.g. images, satellite data etc) so if you are heavy into this you need to use other software.  

## Ask me 3 questions about spatial data in R!